Nitric oxide feedback to ciliary photoreceptor cells orchestrates UV avoidance

### 2023-06-02 Manuscript/Manuscript.Rmd
Ms: edit discussion compartmentalisation section

Our work revealed an essential role for NO-mediated signalling in driving UV/violet avoidance behaviour in larval *Platynereis*.
NO, produced by postsynaptic INNOS interneurons, signals retrogradely to presynaptic cPRCs via NIT-GC1 leading to delayed and sustained cPRC activation.
This late-phase activation of the photoreceptors drives circuit output through projection interneurons and ciliomotor neurons.
In the cPRC circuit, synaptic connectivity alone is thus not sufficient to account for circuit dynamics and behavioural change.

@@

NO is a free radical with a millisecond-to-second half life and thus a limited signalling range.
In neuronal signalling, NOS is often localized to neurites [@Kuntz2017] at close proximity to sGC at synapses [@burette2002; @garthwaite2015a]. In the cPRC circuit, NOS is localised to the input compartment of INNOS neurites in the neurosecretory plexus where cPRC axons synapse on INNOS neurites. NIT-GC1, the target of NO signalling is localised to cPRC projections.
NO-mediated retrograde signalling thus likely occurs in the neurosecretory plexus where NOS- and NIT-GC1-containing projections are in close proximity and where we could detect NO production following UV stimulation.

@@

In the *Platynereis* circuit, our mathematical model indicates that the onset, magnitude and duration of the NO-dependent signal depends on the intensity and duration of the UV/violet stimulus. This suggests that the NO-dependent memory trace also encodes the magnitude and duration of the stimulus. During UV avoidance in the planarian *Schmidtea mediterranea*, neuropeptide signalling has a similar integratory function. Planarians exposed to UV light for >30 sec remain active for extended periods (several minutes) post-stimulation [@bray2023]. If neuropeptide signalling is defective, animals can still respond to UV light, but do not maintain a latent memory state and do not display increased post-stimulus activity.

@@

It is known that complex integrative networks of neurons are involved in UV avoidance behaviour in planaria [@Shettigar2021] and Drosophila [@Imambocus2022].
It has been reported that the sensory neurons required for UV avoidance behaviour in Drosophila converge and arrest neuromodulatory hub neurons, and that compartmentalised circuit organisation and neuropeptide release from the regulatory hubs act as central circuit elements controlling escape responses.
RGW interneurons have been suggested to be important for relaying peptidergic signals due to the distinct anatomy and connectivity of projection neurons in peptidergic networks and their high authority values, and central with many neuronal inputs and outputs network components may function as regulatory hubs that gate the activation of specific networks [@williams2017].
The finding of sGCs in INRGWs from the present single-cell analysis also suggests that they may be regulated by NO.
The specific demarcation of sensory input and output through the use of neurotransmitters, neuropeptides and even gaseous NO may further increase the efficiency of network computation in hub neurons and may lead to a more complex mechanism of 'down' recognition.

>>>>

Our work revealed an essential role for NO-mediated signalling in driving UV/violet avoidance behaviour in larval *Platynereis*.
NO, produced by postsynaptic INNOS interneurons, signals retrogradely to presynaptic cPRCs via NIT-GC1 leading to delayed and sustained cPRC activation.
This late-phase activation of the photoreceptors drives circuit output through projection interneurons and ciliomotor neurons.
In the cPRC circuit, synaptic connectivity alone is thus not sufficient to account for circuit dynamics and behavioural change, as documented in other circuits [@Imambocus2022].

@@

NO is a free radical with a millisecond-to-second half life and thus a limited signalling range.
In neuronal signalling, NOS is often localized to neurites [@Kuntz2017] at close proximity to sGC at synapses [@burette2002; @garthwaite2015a].
In the cPRC circuit, NOS is localised to the dendritic compartment of INNOS neurites in the neurosecretory plexus where cPRC axons synapse on INNOS neurites.
NIT-GC1, the target of NO signalling is localised to cPRC projections.
NO-mediated retrograde signalling thus likely occurs in the neurosecretory plexus where NOS- and NIT-GC1-containing projections are in close proximity and where we could detect NO production following UV stimulation.

Compartmentalised signalling also occurs in peptidergic modulatory system and can be important for selective network activity during specific behaviours. In the UV-avoidance circuit of *Drosophila* larvae, a peptidergic hub neuron Dp7 integrates different sensory inputs occurring at functionally separated dendritic compartments.
The insulin-like peptide 7 (Ilp7) expressed in Dp7 mediates the UV signal from extraocular sensory neurons to motor circuits and is required for acute and sustained UV avoidance [@Imambocus2022].

@@

In the *Platynereis* circuit, our mathematical model indicates that the onset, magnitude and duration of the NO-dependent signal depends on the intensity and duration of the UV/violet stimulus.
This suggests that the NO-dependent memory trace also encodes the magnitude and duration of the stimulus.
During UV avoidance in the planarian *Schmidtea mediterranea*, neuropeptide signalling has a similar integratory function.
Planarians exposed to UV light for \>30 sec remain active for extended periods (several minutes) post-stimulation [@bray2023].
If neuropeptide signalling is defective, animals can still respond to UV light, but do not maintain a latent memory state and do not display increased post-stimulus activity.

@@

It is known that complex integrative networks of neurons are involved in UV avoidance behaviour in planaria [@Shettigar2021]

RGW interneurons have been suggested to be important for relaying peptidergic signals due to the distinct anatomy and connectivity of projection neurons in peptidergic networks and their high authority values, and central with many neuronal inputs and outputs network components may function as regulatory hubs that gate the activation of specific networks [@williams2017].
The finding of sGCs in INRGWs from the present single-cell analysis also suggests that they may be regulated by NO.
The specific demarcation of sensory input and output through the use of neurotransmitters, neuropeptides and even gaseous NO may further increase the efficiency of network computation in hub neurons and may lead to a more complex mechanism of 'down' recognition.
